\chapter{Creating a new directory for source code}

In this section, we describe how setting up a new directory for source
code. In fact, it is very easy to create a new directory, but getting
\Dune build system to recognize the new folder takes some steps which
will be explained in more detail below:

\begin{itemize}
\item Create new directory with content. Quite often an existing
  directory can be used as a base for the new one.
\item Adapt \texttt{Makefile.am} in the new directory and add this
  direcory to the \texttt{SUBDIRS} list of the \texttt{Makefile.am} in
  the directory above the one which you just created.
\item Adapt \texttt{configure.ac} in the \eWoms base directory (the
  directory to which you checked out or unpacked the \eWoms source
  code, probably \texttt{ewoms})
\item Run the command \texttt{automake} in the \eWoms base directory.
\item Re-compile \eWoms by typing \texttt{make} in the \eWoms base
  directory.
\end{itemize}

\noindent In more detail:

\textbf{First} of all, the new directory including all relevant files
needs to be created (see section \ref{tutorial-coupled} and
\ref{tutorial-decoupled} for description of how to define problems).

\textbf{Second}, a new \texttt{Makefile.am} for the new directory
needs to be created. It is good practice to simply copy and modify an
existing file. For example, the file \texttt{tutorial/Makefile.am}
looks as follows:
\begin{verbatim}
# programs just to build when "make check" is used
check_PROGRAMS = tutorial_decoupled tutorial_coupled

noinst_HEADERS= *.hh
EXTRA_DIST = CMakeLists.txt

tutorial_coupleddir = $(datadir)/ewoms/tutorial
tutorial_coupled_SOURCES = tutorial_coupled.cc
tutorial_coupled_DATA = $(tutorial_coupled_SOURCES)

include $(top_srcdir)/am/global-rules
\end{verbatim}

All occurrences of \texttt{tutorial\_coupled} need to be replaced by
the name of the new project, e.g. \texttt{new\_project}. At least if
the name of the source file as well as the name of the new project are
\texttt{new\_project}.

\textbf{Third}: In the parent directory of the directory which you
created, there is also a \texttt{Makefile.am}. In this file the
sub-directories are listed. As you introduced a new sub-directory, it
needs to be included there. In this case, the name of the new
directory is \texttt{new\_project}.

\begin{verbatim}
...
SUBDIRS = doc ewoms m4 test tutorial new_project
...
\end{verbatim}

\textbf{Fourth}: In \eWoms base directory there is a file
\texttt{configure.ac}. In this file, the respective Makefiles are
listed. After a line reading

 \texttt{AC\_CONFIG\_FILES([Makefile} 

 \noindent a line, declaring a new Makefile, needs to be added. The
 Makefile itself will be generated automatically from the
 \texttt{Makefile.am} file which you just generated. For keeping track
 of the included files, it is adviced to insert new Makefiles in
 alphabetical order. The new line could read:
 \texttt{new\_project/Makefile}

 \textbf{Fifth}: Run the command \texttt{automake} in the \eWoms base
 directory.

 \textbf{Sixth}: Re-compile \eWoms as described in Section \ref{install}.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "ewoms-handbook"
%%% End: 




