\chapter[Tutorial]{Tutorial}\label{chp:tutorial}

\eWoms provides two sorts of models: Models which use a fully-implicit
discretization in space and time and models that use a
semi-implicit space and an explicit time discretization.

The semi-implicit models work by first implicitly solving a mass
balance equation for all conservation quaintites combined. This
procedire yields a pressure field. Then, the conserved quantities are
transported using an explicit post-processing procedure based on of
the pressure field obtained in the implicit step. These models are
usually refered to be based on the IMPES approach (\textbf{i}mplicit
\textbf{p}ressure \textbf{e}xplicit \textbf{s}aturation) or, more
generally, on the IMPET approach (\textbf{i}mplicit \textbf{p}ressure
\textbf{e}xplicit \textbf{t}ransport).

In contrast, the fully-implicit models describe the conservation
quantities of a flow system as a system of strongly coupled partial
differential equations which is directly using a non-linear
solver. Physically, these conservation quantities are mass, momentum
and energy; Although the momentum is usually not explicitly conserved
in the context of flow models for porous media.

In section \ref{box} a short introduction to the vertex centered
finite volume scheme (VCFV or box method) used by \eWoms as the
spatial discretization of the fully-implicit models is given. The
decoupled models usually employ a cell-centered finite volume scheme
which is quite similar to the VCFV discretization, except that this
scheme does not require to construct a dual grid centered around the
primary grid's vertices.

This chapter, first introduce how to solve flow problems using a
fully-implicit model (section \ref{tutorial-coupled}) and then how to
solve the same problem using a semi-implicit model (section
\ref{tutorial-decoupled}). Being simple but representative case, an
isothermal two-phase problem (i.e. two fluid phases, one solid phase)
will be considered. The source code of these tutorials is shipped with
the \eWoms source package and can be found in the \texttt{tutorial}
directory.

\input{tutorial-coupled}
\input{tutorial-decoupled}
%\input{tutorial-newmodel}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "ewoms-handbook"
%%% End: 
